---
title: "Introduction to Tidy Network Analysis and Visualization"
author: "Filippo Chiarello"
output:
  beamer_presentation:
    incremental: yes
  ioslides_presentation:
    widescreen: yes
    css: style.css
    incremental: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE)
```

## Overview

* a **network**  is a collection of _vertices joined by edges_ 
* vertices and edges are also called _nodes and links_ in computer science and _actors and ties_ in sociology 
* in mathematics, a network is called **graph** and it is typically represented as a square matrix 
* given a graph $G$ with $n$ nodes numbered from $1$ to $n$, the **adjacency matrix** $A = (a_{i,j})$ of $G$ is a square $n \times n$ matrix such that $a_{i,j} = 1$ if there exists an edge joining nodes $i$ and $j$, and $a_{i,j} = 0$ otherwise 

## Undirected graphs

* a graph is **undirected** if edges have no direction: if there is an edge from $i$ to $j$, then there is also an edge from $j$ to $i$ 
* this means that the adjacency matrix of an undirected graph is **symmetric**: $a_{i,j} = a_{j,i}$ for every pair $i,j$ (or $A = A^T$, where $A^T$ is the transpose of $A$)
* if there is and edge between $i$ and $j$, then $i$ and $j$ are said to be **neighbors** 
* the neighbors of node $i$ are the $1$ of row (or column) $i$ of $A$

## Undirected graphs

```{r, echo= FALSE}
library(igraph)
edges = c(1,2, 1,5, 2,3, 2,4, 3,4, 3,5, 3,6)
ug = graph(edges, directed=FALSE)
coords = layout_with_fr(ug)
plot(ug, layout=coords,
     vertex.size = 20, vertex.color = "lightblue", 
     edge.width = 3, edge.color = "black") 

(A = as_adjacency_matrix(ug, sparse=FALSE))


```

## Directed graphs

* a graph is **directed** if edges have a direction: if there is an edge from $i$ to $j$, then there might be or not the inverse edge
* this means that the adjacency matrix of an directed graph is not necessarily symmetric
* if there is and edge from $i$ to $j$, then $i$ is a **predecessor** of $j$ and $j$ is a **successor** of $i$ 
* the predecessors of node $i$ are the $1$ on column $i$ of $A$ and the successors of node $i$ are the $1$ on row $i$ of $A$
* **self-loops** or self-edges are edges $(i,i)$ from a node $i$ to itself: they correspond to diagonal entries in the adjacency matrix 

## Directed graphs

```{r , echo = FALSE}
edges = c(1,3, 1,4, 2,5, 3,2, 4,4, 4,5, 5,3, 5,6, 5,4)
dg = graph(edges, directed=TRUE)
coords = layout_with_fr(dg)
plot(dg, layout=coords,
     vertex.size = 20, vertex.color = "lightblue", 
     edge.width = 3, edge.color = "black") 

(A = as_adjacency_matrix(dg, sparse=FALSE))
```

## Unweighted and weighted graphs

In some situations it is useful to represent edges as having a strength, weight, or value to them: 

1. in the Internet edges might have weights representing the amount of data flowing along them or their bandwidth
1. in a food web predator-prey interactions might have weights measuring total energy flow between prey and predator
1. in a social network connections might have weights representing the sign and intensity of the relationship: positive weights denote friendship and negative ones represent animosity

- _Any SCI related examples?_

## Unweighted and weighted graphs

```{r, echo = FALSE}
edges = c(1,2, 1,5, 2,3, 2,4, 3,4, 3,5, 3,6)
wg = graph(edges, directed=FALSE)
E(wg)$weight = c(1, 5, 2, 8, 4, 5, 6)
coords = layout_with_fr(wg)
plot(wg, layout=coords,
     vertex.size = 20, vertex.color = "lightblue", 
     edge.width = 1, edge.color = "grey", edge.label = E(wg)$weight)
     
(A = as_adjacency_matrix(wg, sparse=FALSE, attr = "weight"))
```

## Bipartite graphs


* a **bipartite graph** is a graph where there are two kinds of vertices, and edges run from nodes of different types only 
* any network in which the vertices are connected together by common membership of groups of some kind can be represented in this way 
* in sociology such networks are called **affiliation networks**: for instance, scientists coauthoring papers or film actors appearing together in films
* a bipartite network is typically represented with an **incidence matrix**: if $n$ is the number of actors in the network and $g$ is the number of groups, then the incidence matrix $B$ is a $g \times n$ matrix having elements $B_{i,j} = 1$ if group $i$ contains participant $j$ and $B_{i,j} = 0$ otherwise


## Bipartite graphs

```{r, echo = FALSE}
types = c(rep(TRUE,7), rep(FALSE,4))
edges = c(8,1, 8,2, 8,3, 9,2, 9,3, 9,4, 9,5, 10,4, 10,6, 11,5, 11,6, 11,7)
bg = make_bipartite_graph(types, edges, directed=FALSE)
lay = layout.bipartite(bg)
plot(bg, layout=lay[,2:1], 
     vertex.color=c("skyblue","green")[V(bg)$type + 1], 
     vertex.size = 20)

(B = as_incidence_matrix(bg))

```

## Tidy network data?

* there is a discrepancy between network data and the tidy data idea, in that network data cannot in any meaningful way be encoded as a single tidy data frame 
* on the other hand, both node and edge data by itself fits very well within the tidy concept as each node and edge is, in a sense, a single observation 
* thus, a close approximation of tidyness for network data is two tidy data frames, one describing the node data and one describing the edge data

## tidygraph

* [tidygraph](https://cran.r-project.org/web/packages/tidygraph/index.html) is an entry into the tidyverse that provides a tidy framework for network (graph) data 
* tidygraph provides an approach to manipulate node and edge data frames using the interface defined in the **dplyr** package
* moreover it provides tidy interfaces to a lot of common graph algorithms, including **igraph** network analysis toolkit
* it is developed by [Thomas Lin Pedersen](https://twitter.com/thomasp85)

## ggraph

* [ggraph](https://cran.r-project.org/web/packages/ggraph/index.html) is an extension of ggplot2 that implements a visualization grammar for network data 
* it provides a huge variety of geoms for drawing nodes and edges, along with an assortment of layouts making it possible to produce a very wide range of network visualization types
* while **tidygraph** provides a manipulation and analysis grammar for network data (like **dplyr** for tabular data), **ggraph** offers a visualization grammar (like **ggplot** for tabular data)
* it is developed by [Thomas Lin Pedersen](https://twitter.com/thomasp85)

## A full example: dplyr, tidygraph and ggraph


```{r}
library(dplyr)
library(ggraph)
library(tidygraph)

# setting theme_graph 
set_graph_style()

# a graph of highschool friendships
head(highschool)

# create the graph and add popularity
graph <- as_tbl_graph(highschool) %>% 
    mutate(Popularity = centrality_degree(mode = "in"))

# plot using ggraph
ggraph(graph, layout = "kk") + 
    geom_edge_link() + 
    geom_node_point(aes(size = Popularity)) + 
    facet_edges(~year) + 
    theme_graph(foreground = "steelblue", fg_text_colour = "white")

```

## Read the graph with tidygraph

Let's read a dolphin network:

> 1. a set of nodes representing dolphins ([dolphin_nodes.csv](data/dolphin_nodes.csv))  
> 2. a set of edges representing ties among dolphins ([dolphin_edges.csv](data/dolphin_edges.csv))

```{r}
library(readr)

nodes <-  read_csv("data/dolphin_nodes.csv")
edges <-  read_csv("data/dolphin_edges.csv")

nodes
edges
```

## About the Data

David Lusseau, a researcher at the University of Aberdeen, observed the group of dolphins of Doubtful Sound. Every time a school of dolphins was encountered in the fjord between 1995 and 2001, each adult member of the school was photographed and identified from natural markings on the dorsal fin. This information was utilised to determine how often two individuals were seen together. Read the [full story](https://medium.com/@cento_otto/a-social-network-of-dolphins-c073f3b8502d).


## Tidygraph data structure

Package tidygraph represents the graph as a pair of data frames:

* a data frame for nodes containing information about the nodes in the graph 
* a data frame for edges containing information about the edges in the graph. The terminal nodes of each edge must either be encoded in a `to` and `from` column, or in the two first columns, as integers. These integers refer to nodes index.

---

```{r}

# add edge type
edges <-  edges %>% 
  mutate(type = sample(c("love", "friendship"), 
                       nrow(edges), 
                       replace = TRUE) )

# make a tidy graph
dolphin <-  tbl_graph(nodes = nodes, edges = edges, directed = FALSE)
dolphin

```

## How to pull back nodes and edges information?

**Pipe friendly** version

```{r eval=FALSE}

# extract node and edge data frames from the graph
dolphin %>% 
  as.list()

# extract node data frame from the graph

dolphin %>% 
  activate(nodes) %>% 
  as_tibble()

# extract edge data frame from the graph
dolphin %>% 
  activate(edges) %>% 
  as_tibble()

```

## Tidygraph visualisation with ggraph

Ggraph builds upon three core concepts that are quite easy to understand:

* the **layout** defines how nodes are placed on the plot. ggraph has access to all layout functions available in [igraph](https://igraph.org/) and much more
* the **nodes** are the connected entities in the graph structure. These can be plotted using the `geom_node_*()` family of geoms
* the **edges** are the connections between the entities in the graph structure. These can be visualized using the `geom_edge_*()` family of geoms 

## ggraph basics

```{r}
# setting theme_graph 
set_graph_style()

# basic node plot
ggraph(dolphin) + 
  geom_node_point()

# basic graph plot
ggraph(dolphin) + 
  geom_edge_link() + 
  geom_node_point()

# plot edge type
ggraph(dolphin) + 
  geom_edge_link(aes(color = type)) + 
  geom_node_point()

# plot node sex
ggraph(dolphin) + 
  geom_edge_link(aes(color = type)) + 
  geom_node_point(aes(shape = sex))

# plot node name
ggraph(dolphin) + 
  geom_edge_link() + 
  geom_node_point() + 
  geom_node_text(aes(label = name), repel=TRUE)

```

## Faceting

Faceting allows to create sub-plots according to the values of a qualitative attribute on nodes or edges.

```{r}
# setting theme_graph 
set_graph_style()


# facet edges by type
ggraph(dolphin) + 
  geom_edge_link(aes(color = type)) + 
  geom_node_point() +
  facet_edges(~type)


# facet both nodes and edges
ggraph(dolphin) + 
  geom_edge_link() + 
  geom_node_point() +
  facet_graph(type~sex) + 
  th_foreground(border = TRUE)
```

## Directed graphs

```{r}
# setting theme_graph 
set_graph_style()

# direcred graphs
package <-  tibble(
  name = c("igraph", "ggraph", "dplyr", "ggplot", "tidygraph")
)

tie <-  tibble(
  from = c("igraph", "ggplot", "igraph", "dplyr", "ggraph"),
  to =   c("tidygraph", "ggraph", "tidygraph", "tidygraph", "tidygraph")
)

tidy <-  tbl_graph(nodes = package, edges = tie, directed = TRUE)


# use arrows for directions
ggraph(tidy, layout = "graphopt") + 
    geom_edge_link(aes(start_cap = label_rect(node1.name), 
                       end_cap = label_rect(node2.name)), 
                   arrow = arrow(type = "closed", 
                                 length = unit(3, "mm"))) + 
    geom_node_text(aes(label = name))

# use edge alpha to indicate direction, 
# direction is from lighter to darker node
ggraph(tidy, layout = 'graphopt') + 
    geom_edge_link(aes(start_cap = label_rect(node1.name), 
                       end_cap = label_rect(node2.name), 
                       alpha = stat(index)), 
                   show.legend = FALSE) + 
    geom_node_text(aes(label = name))
```

## Hierarchical layouts

```{r}
# setting theme_graph 
set_graph_style()

# This dataset contains the graph that describes the class 
# hierarchy for the Flare visualization library.
head(flare$vertices)
head(flare$edges)

# flare class hierarchy
graph = tbl_graph(edges = flare$edges, nodes = flare$vertices)

# dendrogram
ggraph(graph, layout = "dendrogram") + 
  geom_edge_diagonal()

# circular dendrogram
ggraph(graph, layout = "dendrogram", circular = TRUE) + 
  geom_edge_diagonal() + 
  geom_node_point(aes(filter = leaf)) + 
  coord_fixed()

# rectangular tree map
ggraph(graph, layout = "treemap", weight = size) + 
  geom_node_tile(aes(fill = depth), size = 0.25)

# circular tree map
ggraph(graph, layout = "circlepack", weight = size) + 
  geom_node_circle(aes(fill = depth), size = 0.25, n = 50) + 
  coord_fixed()

# icicle
ggraph(graph, layout = "partition") + 
  geom_node_tile(aes(y = -y, fill = depth))

# sunburst (circular icicle)
ggraph(graph, layout = "partition", circular = TRUE) +
  geom_node_arc_bar(aes(fill = depth)) +
  coord_fixed()
```


## Network analysis with tidygraph

* the data frame graph representation can be easily augmented with metrics computed on the graph
* before computing a metric on nodes or edges use the `activate()` function to activate either node or edge data frames
* use dplyr verbs filter, arrange and mutate to manipulate the graph

## Network analysis with tidygraph

```{r}
dolphin <-  dolphin %>% 
  activate(nodes) %>% 
  mutate(degree = centrality_degree()) %>% 
  filter(degree > 0) %>% 
  arrange(-degree) %>% 
  activate(edges) %>% 
  mutate(betweenness = centrality_edge_betweenness()) %>% 
  arrange(-betweenness)

dolphin
```

## Analyse and visualize network: centrality

Packages tidygraph and ggraph can be pipelined to perform analysis and visualization tasks in one go.

```{r}
# setting theme_graph 
set_graph_style()


dolphin %>% 
  activate(nodes) %>%
  mutate(pagerank = centrality_pagerank()) %>%
  activate(edges) %>%
  mutate(betweenness = centrality_edge_betweenness()) %>%
  ggraph() +
  geom_edge_link(aes(alpha = betweenness)) +
  geom_node_point(aes(size = pagerank, colour = pagerank)) + 
  # discrete colour legend
  scale_color_gradient(guide = "legend")

# or even less typing
ggraph(dolphin) +
  geom_edge_link(aes(alpha = centrality_edge_betweenness())) +
  geom_node_point(aes(size = centrality_pagerank(), 
                      colour = centrality_pagerank())) + 
  scale_color_gradient(guide = "legend")
```

## Analyse and visualize network: communities

```{r}
# setting theme_graph 
set_graph_style()


# visualize communities of nodes
dolphin %>% 
  activate(nodes) %>%
  mutate(community = as.factor(group_louvain())) %>% 
  ggraph() + 
  geom_edge_link() + 
  geom_node_point(aes(colour = community), size = 5)

```

## Analyse and visualize network: graph metrics

Compute degree, closeness, betweenness and PageRank on the network. Are top-ranked dolphins male or female? 

```{r}

dolphin <- dolphin %>% 
  activate(nodes) %>% 
  mutate(degree = centrality_degree(), 
         pagerank = centrality_pagerank(), 
         closeness = centrality_closeness(),
         betweenness = centrality_betweenness())

actors <-  as.list(dolphin)$nodes

actors %>% 
  group_by(sex) %>% 
  summarise(avg_degree = mean(degree),
            avg_pagerank = mean(pagerank),
            avg_closeness = mean(closeness),
            avg_betweenness = mean(betweenness))


```

## Analyse and visualize network: graph metrics

Study the correlation among the four centrality measures.

```{r}

M <-  cbind(degree = actors$degree,
          pagerank = actors$pagerank,
          closeness = actors$closeness,
          betweenness = actors$betweenness)


(corM <-  cor(M))
corrplot(corM, method = "ellipse")

```


# Thanks for your attetion!

Filippo Chiarello


